---
title: "Classical Statistical Approach to GxE analysis"
author: 
  - name: "Moshood Agba Bakare"
    affiliation: "JeanLuc's Lab Cornell University"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{amsmath}
output:
  bookdown::html_document2:
    fig_caption: true
    number_sections: true
    df_print: paged
    mathjax: "default"
    highlight: kate
    toc: true
    toc_depth: 3
    #css: "my_style.css"
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
# Configuring global parameters
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error=FALSE, warning = FALSE,
                      fig.align = "center", fig.height = 6, fig.width = 6,
                      class.source = "lineAnchors")
```


```{r, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#=======================================================================
# Packages
#=======================================================================
if (!require("pacman")) {
  install.packages("pacman")
}
# Include all packages here
pacman::p_load(
  plyr,
  ggplot2,
  ggrepel,
  factoextra,
  dplyr,
  tidyverse,
  tidyr,
  tibble,
  statgenGxE,
  asreml,
  RColorBrewer,
  kableExtra, 
  prettycode,
  formattable
)
prettycode::prettycode() 
```


# Read combined GxE data in csv format into R 
The combined GxE data is imported into R and tranposed from wide to narrow format for easy visualization of key variables of interest.

```{r,import}
# Import the combined raw data in wide format
phenoDat <- read.csv(file="data/combined_trial.csv", header=TRUE)


# Subset key variable and transpose the imported data in wide format to a narrow format
narrowPhenoDat <- phenoDat %>%
  dplyr::select(loc, year, env, trial, rep, gen,prop_hav, fyld, dyld, hi, tyld, dm) %>%
  gather(key = trait, value = y, -c(loc, year, env, trial, rep, gen, prop_hav))

# write the narrow format combined GxE data file to a working directory data
write.csv(narrowPhenoDat,file="output/narrowPhenoDat.csv", row.names=FALSE)

# coerce variable to factors
narrowPhenoDat$trial <- as.factor(narrowPhenoDat$trial)
narrowPhenoDat$loc <- as.factor(narrowPhenoDat$loc)
narrowPhenoDat$year <- as.factor(narrowPhenoDat$year)
narrowPhenoDat$env <- as.factor(narrowPhenoDat$env)
narrowPhenoDat$trait <- as.factor(narrowPhenoDat$trait)

narrowPhenoDat$year <- factor(narrowPhenoDat$year, levels = c("2018", "2019", "2020"), labels = c("2017-2018", "2018-2019", "2019-2020"))

```


# Data visualization
This is a boxplot showing the distribution of key traits of interest across environments i.e. location and year combination

```{r,boxplot_env_trait}
boxplot_env_trait <- ggplot(data = narrowPhenoDat , aes(x=loc,y=y, fill=trait)) +
    geom_boxplot( stat = "boxplot",  outlier.colour = "red", outlier.shape = 16, outlier.size = 0.5, na.rm=TRUE) +
  stat_summary(fun=mean, colour="blue", geom="point", na.rm = TRUE) + 
  labs(x= "Location", y= "Response values") + theme_bw() +
  theme(axis.title = element_text(colour="black",face="bold", size=12),
         plot.title = element_text(hjust = 0.5,lineheight=.5,colour="black",face="bold", size=12),
         axis.text = element_text(face="bold", size=7), axis.text.x = element_text(angle = 90, hjust = 1),legend.position="none")+
    facet_grid(trait~year, scales = "free")

# save the plot to a file
ggsave("output/boxplot_env_trait.jpeg",height=5, width=8, units="in", dpi=300)
print(boxplot_env_trait)

```


# A diagnostic plot I
This is a diagnostics plot of CV vs broad-sense heritability of each trial. The input file to generate this plot is a summary statistics of raw data resulting from using SAS application.

```{r, CV_vs_H2, results='hide'}
# read summary statistics output file
stats.output <- read.csv(file="output/raw_pheno_trial_summary.csv", header=T)

# filter the trait of interest from the summary output file
stats.output <- subset(stats.output, trait=="fyld") # subset for fresh root yield trait

cv_vs_H2 <- ggplot(data = stats.output,
       aes(x = H2,
           y = cv)) + 
  geom_point(color = ifelse(stats.output$cv > 40.5 | stats.output$H2 < 0.05, "#E41A1C", "#4DAF4A"),size = 1.7,
             alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_text_repel(aes(x = H2,
                      y = cv,
                      label = trial),
                  size = 2.5, segment.colour = NA) +
  labs(x = "Heritability",
       y = "CV (%)") +
 theme_bw() +
   scale_x_continuous(breaks = seq(from = 0.0, to = 0.70, by = 0.1))
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(size = 7, hjust = 0.5),
        axis.text.y = element_text(size = 7, hjust = 0.5, angle = 90),
       axis.title.y = element_text(size = 10, face = "bold"),
       axis.title.x = element_text(size = 10, face = "bold"),
       legend.background = element_rect(fill = "NA", color = "black", size = 0.5),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 10, face = "bold"),
       legend.position = "right"
 )

# save the plot to a file
ggsave("output/cv_vs_H2.jpeg",height=5, width=8, units="in", dpi=300)

# Print the heatmap
print(cv_vs_H2)
```


# A diagnostic plot II
This is a  diagnostics plot of experimental accuracy vs broad-sense heritability of the trials. The input file to generate this plot is a summary statistics of raw data resulting from using SAS application.

```{r,Ac_vs_H2}

# filter the trait of interest from the summary output file
Ac_H2_output <- subset(stats.output, trait=="fyld", select= c("trial","trait","H2","Accr"))

Ac_vs_H2 <- ggplot(data = Ac_H2_output,
       aes(x = H2,
           y = Accr)) +
  geom_point(color = ifelse(Ac_H2_output$Accr < 0.40 | Ac_H2_output$H2 < 0.05, "red", "green"),size = 1.7,
             alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_text_repel(aes(x = H2,
                      y = Accr,
                      label = trial),
                  size = 2.5, segment.colour = NA) +
  labs(x = "Heritability",
       y = "Experimental accuracy") +
 theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(size = 7, hjust = 0.5),
        axis.text.y = element_text(size = 7, hjust = 0.5, angle = 90),
       axis.title.y = element_text(size = 10, face = "bold"),
       axis.title.x = element_text(size = 10, face = "bold"),
       legend.background = element_rect(fill = "NA", color = "black", size = 0.5),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 10, face = "bold"),
       legend.position = "right"
 )


# output the plot to a file
ggsave("output/Ac_vs_H2.jpeg",height=5, width=8, units="in", dpi=300)

print(Ac_vs_H2) # output the plot
```


# A line graph showing the average minimum and maximum temperature, total precipitation, and average relative humidity across the 20 testing environments.

```{r}
weatherDat <- read.csv(file = "data/weatherData.csv", header=TRUE)

weather <- weatherDat %>%
  group_by(ENV,AGROZONE) %>%
  dplyr::summarize(TMIN=mean(TMIN, na.rm=TRUE),
                   TMAX=mean(TMAX, na.rm=TRUE),
                   PRECIPITATION = sum(PRECIPITATION, na.rm = TRUE),
                   RHMEAN = mean(RHMEAN, na.rm=TRUE))%>%
  gather(key = parameter, value = y, -c(ENV,AGROZONE)) %>%
  mutate(parameter=as.factor(parameter))

weather$parameter <- factor(weather$parameter, levels = c("TMIN",
                                                          "TMAX",
                                                          "PRECIPITATION","RHMEAN"), 
                  labels = c("Min Temp (Cel.)",
                             "Max Temp (Cel.)",
                             "Precipitation (mm)", "Mean RH (%)"))

weatherDataPlot <- ggplot(data=weather,aes(x=ENV,y=y,group=1, color=AGROZONE))+geom_line()+geom_point()+
  labs(x="Environment", y="Estimates") +
  theme_bw() +
  theme(axis.title=element_text(colour="black",face="bold", size=12),
        axis.text = element_text(size=7),
        axis.text.x = element_text(angle = 90, size=10,hjust = 1),
        axis.text.y = element_text(size=10),
        strip.text = element_text(size = 6, color = "black",lineheight=5.0))+
  guides(col=guide_legend("Agro-ecological zones"))+
  facet_grid(rows=vars(parameter),scales = "free")

# output the plot to a file
ggsave("output/weatherDataPlot.jpeg",height=4.2, width=6.5, units="in", dpi=300)
print(weatherDataPlot)
```



# Boxplot showing distribution of traits for parameter estimate

Boxplot showing distribution of fresh root yield (fyld t/ha), dry matter content (dmc %), dry yield (dyld t/ha), harvest index (hi), and top yield (tyld t/ha) for derived parameters such as broad-sense heritability, coeficient of variation, experimental accuracy, and residual variance across 17 trials.

```{r,boxplotParameterEstimate}
trialSummStats <- read.csv("output/raw_pheno_trial17_summary.csv", header = T)

summStats <- trialSummStats %>%
  dplyr::select(trial,trait,H2,cv,Accr,res_var) %>%
  gather(key = parm, value = estm, -c(trial, trait))

summStats$trait <- as.factor(summStats$trait)
summStats$parm <- as.factor(summStats$parm)

# reorder level of factors
summStats$parm <- factor(summStats$parm, levels = c("H2", "cv", "Accr","res_var"), labels=c("Broad-sense heritability","Coefficient of correlation",
"Experimental accuracy","Residual variance"))


summStats$trait <- factor(summStats$trait, levels = c("fyld", "dm", 
                                                      "dyld","hi","tyld"))

boxplot_trait_parm <- ggplot(data = summStats , 
                             aes(x=trait,y=estm, fill = "grey30", 
                                 alpha = 0.5)) +
  geom_boxplot(stat = "boxplot",  
                 outlier.colour = "grey30", 
                 outlier.alpha = 0.5, 
                 outlier.shape = 16, 
                 outlier.size = 0.5, 
                 na.rm=TRUE) +
  stat_summary(fun=mean, 
               colour="blue", 
               geom="point", na.rm = TRUE) +
  
  geom_jitter(width = 0.2, 
              color = "black", 
              alpha = 0.5) +
  labs(y="Parameter estimate") + 
  theme_bw() +
  theme(axis.title=element_text(colour="black",face="bold", size=12),
         axis.text = element_text(size=10),
        legend.position="none") +
  facet_wrap(vars(parm), ncol = 4, scales="free_y")
  

ggsave("output/boxplot_trait_parm.jpeg",height=4.2, width=6.5, units="in", dpi=300)
print(boxplot_trait_parm)

```


# Percentage of variation attributed to factors in a classical random anova model

This is a plot showing the percentage of total variation attributed to environment, genotype, genotype by environment, rep nested within environment, and residual term from fitting joint random effect model

```{r}
anovaParms <- read.csv("combined_GE_parameter.csv", header = T)

# coerce trait to a factor and reorder
anovaParms$Effect <- as.factor(anovaParms$Effect)
anovaParms$trait <- as.factor(anovaParms$trait)
anovaParms$trait <- factor(anovaParms$trait,
                           levels = c("fyld", "dmc","dyld","hi","tyld"),
                           labels = c("Fresh root yield", "Dry matter content","Dry yield","Harvest index","Top yield"))

                    
anovaParmsPlot<- ggplot(data=anovaParms, aes(x=trait,y=percent_var,fill=Effect))+
  geom_bar(position="dodge",stat="identity",color="grey30") + 
  coord_flip() +
  labs(y="Percentage of total variance explained") +
  theme_bw() +
  theme(axis.title = element_text(colour="black",face="bold", size=12),
         plot.title = element_text(hjust = 0.5,lineheight=.5,colour="black",face="bold", size=12),
         axis.text = element_text(face="bold", size=10), axis.text.x = element_text(angle = 0, hjust = 1))+
  scale_fill_grey(start = 0, end = .9)

anovaParmsPlot
ggsave("output/anovaParmsPlot.jpeg",height=4.2, width=6.5, units="in", dpi=300)
print(anovaParmsPlot)
```


# Fitting Finlay Wilkinson (FW) model

This chunk called a defined function fwModel to fit FW model for a trait of interest.The script reads two-way GxE adjusted means output from  fitting linear mixed  using SAS application.

From this point, the analysis was based on 17 trials after data quality control where 3 trials are excluded as shown in red data point in diagnostic plots.

Note: The function can be invoked to fit FW model for any trait of interest within the input data file.

```{r, fwModel}
# import the two-way GxE table ouput resulting from using SAS application
two_wayGxE <- read.csv(file="output/gen_env_trait.csv", header=T)

# Standardize the trait to mean of zero and unit variance be able to compare MSE across the traits

fwScaleDat <- two_wayGxE %>%
  mutate_at(c("dyld","fyld","hi","tyld","dm"),~(scale(.)%>% as.vector))

## Create a trial data TD object
setATD <- statgenSTA::createTD(data = fwScaleDat, genotype = "gen", trial = "env")

# Invoke a defined function fwMpdel from source script to fit FW model
source("code/fwModel.R")

# Example: Invoke the defined function to fit FW model for fresh root yield (fyld)
fwOutput <- fwModel(traitName = "fyld")
print(fwOutput)
```


# Boxplot of means square error (MSE) and slope from FW  model

This Boxplot showing the distribution of the parameter estimates - slope and MSE from Finlay Wilkinson model

```{r}
fwSensMse <- read.csv(file="fwScaleTraitSens.csv", header = T)

fwParmEst <- fwSensMse %>%
  dplyr::select(genotype, trait,sens,MSdeviation) %>%
  rename(MSE=MSdeviation)%>%
  gather(key = parm, value = estm, -c(genotype, trait))


# Coerce to a factor
fwParmEst$trait <- as.factor(fwParmEst$trait)
fwParmEst$parm <- as.factor(fwParmEst$parm)

# reorder the factor levels
fwParmEst$trait <- factor(fwParmEst$trait, levels = c("fyld", "dm", 
                                                      "dyld","hi","tyld"))

fwParmEst$parm <- factor(fwParmEst$parm, levels = c("MSE", "sens"), labels=c("Means square error","Slope"))

boxplot_fwParm <- ggplot(data = fwParmEst,
                             aes(x=trait,y=estm, fill = "grey30", 
                                 alpha = 0.5)) +
  geom_boxplot(stat = "boxplot",  
                 outlier.colour = "grey30", 
                 outlier.alpha = 0.5, 
                 outlier.shape = 16, 
                 outlier.size = 0.5, 
                 na.rm=TRUE) +
  stat_summary(fun=mean, 
               colour="blue", 
               geom="point", na.rm = TRUE) +
  
  geom_jitter(width = 0.2, 
              color = "black", 
              alpha = 0.5) +
  labs(y="Parameter estimate") + 
  theme_bw() +
  theme(axis.title=element_text(colour="black",face="bold", size=12),
         axis.text = element_text(size=10),
        legend.position="none",
        strip.text = element_text(size = 12, color = "black")) +
  
  facet_wrap(vars(parm), ncol = 2, scales="free_y")


  ggsave(paste("output/boxplot_fwParm",".jpeg",sep=""),
       plot=boxplot_fwParm,
       height=4.2, width=6.5, units="in", dpi=300)
  
 print(boxplot_fwParm) 
```


# Summary statistics of MSE and slope
This computes the minimum, median, maximum, and variance of means square (MSE) and slope resulting from fitting FW model for the observed traits

```{r}
fwParmEstStats <- fwParmEst %>%
  group_by(parm,trait)%>%
  summarise(minimum=min(estm, na.rm=T),median=median(estm, na.rm=T),
            maximum = max(estm, na.rm=T), variance=var(estm,na.rm=T))
write.csv(fwParmEstStats,file="output/fw_MSE_and_slope_statistics.csv",row.names = FALSE)
```



# Fitting Additive Main Effect and Multiplicative (AMMI) Model
A defined function ammiModel was invoked to fit AMMI model on fresh root yield. The script reads two-way GxE adjusted means output from  fitting linear mixed  using SAS application.

Like FW model, the analysis was based on 17 trials after data quality control where 3 trials are excluded as shown in red data point in diagnostic plots.


```{r, ammiModel}

# Invoke a defined funtion ammiModel from source script to fit AMMI model
source("code/ammiModel.R")

# Example: Invoke the defined function to fit AMMI model for fresh root yield (fyld)
AMMIoutput <- ammiModel(traitName = "fyld")
print(AMMIoutput)
```



# Fit genotype and genotype by environment interaction (GGE) model
A defined function ggeModel was invoked to fit GGE model on fresh root yield. The script reads two-way GxE adjusted means output from fitting linear mixed  using SAS application.

Like other two models, the analysis was based on 17 trials after data quality control where 3 trials are excluded as shown in red data point in diagnostic plots.

```{r, ggeModel}

# Invoke a defined funtion ggeModel from source script to fit GGE model
source("code/ggeModel.R")

# Example: Invoke the defined function to fit GGE model for fresh root yield (fyld)
GGEoutput <- ggeModel(traitName = "fyld")
print(GGEoutput)
```


# Percentage of variation attributed to each factor in linear-bilinear model

This is a Barchart showing percentage of variation attributed to environment, genotype, genotype by environment interaction, and residual terms in FW, AMMI, and GGE models.

```{r}
modelParms <- read.csv("ammi_barGraph.csv", header = T)

# coerce trait to a factor and reorder
modelParms$trait <- as.factor(modelParms$trait)
modelParms$trait <- factor(modelParms$trait,
                           levels = c("fyld", "dmc","dyld","hi","tyld"),
                           labels = c("Fresh root yield","Dry matter content",
                                     "Dry yield",  "Harvest index","Top yield"))

modelParmsPlot <- ggplot(data=modelParms, aes(x=trait,y=TSS,fill=factor(Model))) +
  geom_bar(position="dodge",stat="identity") + 
  coord_flip() +
  labs(y="Percentage of total variation explained", fill="Model") +
  theme_bw() +
  theme(axis.title = element_text(colour="black",face="bold", size=12),
         plot.title = element_text(hjust = 0.5,lineheight=.5,colour="black",face="bold", size=12),
         axis.text = element_text(face="bold", size=8), axis.text.x = element_text(angle = 0, hjust = 1))+
  scale_fill_grey(start = 0, end = .9)+
  facet_grid(.~source, scales = "free") + theme(strip.text = element_text(
    size = 7,face = "bold", color = "black"))
print(modelParmsPlot)

ggsave("output/parmEstimatePlot.jpeg",height=4.2, width=6.5, units="in", dpi=300)
```



# Compute univariate stability statistic (Cultivar superiority index)

This section assesses the consistency of genotype stability performance by exploring
Cultivar superiority index as as univariate stability statistics.

```{r,univariateStabilityMeasure}
# Create a trial data TD object from two-way data structure.
setATD <- statgenSTA::createTD(data = two_wayGxE, genotype = "gen", trial = "env")

# compute Cultivar-superiority measure for fyld

fyldgeStab <- gxeStability(TD = setATD, trait = "fyld")

# extract coefficient of stability value
fyldSupIndex <- fyldgeStab$superiority
colnames(fyldSupIndex)[2:3] <- c("fyld","fyld_SupIndexValue")


# compute Cultivar-superiority measure for dm
dmgeStab <- gxeStability(TD = setATD, trait = "dm")

# extract coefficient of stability value
dmSupIndex <- dmgeStab$superiority
colnames(dmSupIndex)[2:3] <- c("dm","dm_SupIndexValue")


### compute Cultivar-superiority measure for dyld
dyldgeStab <- gxeStability(TD = setATD, trait = "dyld")

# extract coefficient of stability value
dyldSupIndex <- dyldgeStab$superiority
colnames(dyldSupIndex)[2:3] <- c("dyld","dyld_SupIndexValue")

# Merge the computed stability index
genStab <- merge(fyldSupIndex, dmSupIndex) %>%
              merge(dyldSupIndex)


# rank ascending order of superiority index for fyld, dm, and dyld
  genStab <- arrange(genStab,fyld_SupIndexValue) %>%
    mutate(fyld_rank=1:nrow(genStab))

  genStab <- arrange(genStab,dm_SupIndexValue) %>%
    mutate(dm_rank=1:nrow(genStab))

  genStab <- arrange(genStab,dyld_SupIndexValue) %>%
    mutate(dyld_rank=1:nrow(genStab)) 


genStab <-  genStab %>%
    select(genotype, fyld, fyld_SupIndexValue, fyld_rank,
           dm, dm_SupIndexValue, dm_rank,
           dyld, dyld_SupIndexValue, dyld_rank)

# print out the result of univariate stability measures
  print(genStab)

# write the output of univariate stability statistics to a file  
write.csv(genStab, file="output/cultivarStabilityIndexOutput.csv",row.names=FALSE) 
```



# scatter plot of cultivar superiority index versus meam of traits
This plot revealed how cultivar superiority index is related to mean of fresh root, dry matter content and dry yield towards identofying high yielding and stable genotypes

```{r,cultivarSupIndexPlot}
fyldCultIndex <- ggplot(data = genStab,
       aes(x = fyld,
           y = fyld_SupIndexValue)) + 
  geom_point(color = ifelse(genStab$fyld_SupIndexValue < 82.0, "#E41A1C", "#4DAF4A"),size = 1.7,
             alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_text_repel(aes(x = fyld,
                      y =fyld_SupIndexValue ,
                      label = genotype),
                  max.overlaps=20,
                  size = 2.5, segment.colour = NA) +
  labs(x = "Mean fresh root yield (t/ha)",
       y = "Cultivar superiority index") +
 theme_bw() +
  theme(axis.title.x = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 10, face = "bold"))

# save the plot to a file
#ggsave("fyldCultIndex.jpeg",height=5, width=8, units="in", dpi=300)

# Print the heatmap
print(fyldCultIndex)


### dry matter content

dmCultIndex <- ggplot(data = genStab,
       aes(x = dm,
           y = dm_SupIndexValue)) + 
  geom_point(color = ifelse(genStab$dm_SupIndexValue < 6.5, "#E41A1C", "#4DAF4A"),size = 1.7,
             alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_text_repel(aes(x = dm,
                      y =dm_SupIndexValue ,
                      label = genotype),
                  max.overlaps=35,
                  size = 2.5, segment.colour = NA) +
  labs(x = "Mean dry matter content (%)",
       y = "Cultivar superiority index") +
 theme_bw() +
  theme(axis.title.x = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 10, face = "bold"))

# save the plot to a file
#ggsave("dmCultIndex.jpeg",height=5, width=8, units="in", dpi=300)

# Print the heatmap
print(dmCultIndex)


### dry yield
  
  dyldCultIndex <- ggplot(data = genStab,
       aes(x = dyld,
           y = dyld_SupIndexValue)) + 
  geom_point(color = ifelse(genStab$dyld_SupIndexValue < 9.5, "#E41A1C", "#4DAF4A"),size = 1.7,
             alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_text_repel(aes(x = dyld,
                      y =dyld_SupIndexValue ,
                      label = genotype),
                  max.overlaps=40,
                  size = 2.5, segment.colour = NA) +
  labs(x = "Mean dry yield (t/ha)",
       y = "Cultivar superiority index") +
 theme_bw() +
  theme(axis.title.x = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 10, face = "bold"))
  
# save the plot to a file
ggsave("output/dyldCultIndex.jpeg",height=5, width=8, units="in", dpi=300)

# Print the heatmap
print(dyldCultIndex)  
```



# Representative of target population of environment(TPE) for fresh yield

This script chunk is determing the number of environments that is likely to represent target population of environment (TPE). This has to be run separately for fyld, dm, and tyld. The output from these 3 traits are merged together to have a plot in the next chunk.

Note: This script chunk for TPE was adapated from Gemenet Dorcus Chepkesis of 
excellent in breeding (EiB).

```{r repTPE, eval=FALSE, echo=FALSE}
pheno <- phenoDat %>% dplyr::select(loc, year, env,trial,
                                    gen,rep,prop_hav,fyld, dm,tyld) %>% 
  dplyr::filter(!trial %in% c("18UYT36setAKN", "19UYT36setAZA", "19UYT36setAMK")) %>% data.frame
           
pheno$rep  <- as.factor(pheno$rep)
pheno$loc  <- as.factor(pheno$loc)
pheno$year <- as.factor(pheno$year)
pheno$env  <- as.factor(pheno$env)
pheno$gen  <- as.factor(pheno$gen)

# Combined analysis model fresh root yield
modfit  <-  asreml(fixed=fyld~env + prop_hav,
        random= ~ at(env):rep + gen + gen:env,
	residual=~dsum(~units|env),
        na.action=na.method(y='include',x='include'),
        data=pheno,maxit=30,workspace = 250e6, pworkspace=250e6,trace = FALSE)


#summary(modfit)$varcomp

preds <- predict(modfit, classify = "gen")$pvals

BLUPlist_All <- preds %>% select(gen,predicted.value) %>%
rename(AllEnvBlup=predicted.value) %>% data.frame

#cat("\n BLUPs across all environments","\n")  
#print(BLUPlist_All)

 
# dataframe to store the result
BlupOutput <- expand.grid(gen = levels(pheno$gen))
heritabOutput <- expand.grid(env = levels(pheno$env))
heritabOutput$H2_Cullis <- NA
heritabOutput$Corr <- NA

# Fit environment-specific model - random genetic effect

for (i in 1:length(levels(pheno$env))){ # loop each trial
  
  	subdat <- subset(pheno,env==unique(levels(pheno$env))[i])
  	modfit <- asreml(fixed = fyld~rep + prop_hav,
			random = ~ gen,
			residual=~id(units),
			na.action=na.method(y='include',x='include'),
			data  = subdat,trace=FALSE)

	# extract out the predicted value from each trial and correlate with all env blups
	PredVal <- predict(modfit, classify = "gen")$pvals
	envPredVal <- PredVal %>% select(gen,predicted.value) %>%
	rename(envBlup=predicted.value) %>% data.frame
	mergePredVal  <- merge(BLUPlist_All, envPredVal, by="gen",all=TRUE)
	heritabOutput$Corr[i] <- with(mergePredVal, cor(AllEnvBlup,envBlup, use="complete.obs"))
	
	colnames(envPredVal)[2] <- levels(pheno$env)[i]
	
	# extract the blups for each environment
	BlupOutput <- merge(BlupOutput, envPredVal, by="gen",all=TRUE)
 
	vc.g <- summary(modfit)$varcomp['gen','component']
	# Mean variance of a difference of two genotypic BLUPs
      	vdBLUP.mat <- predict(modfit, classify="gen", only="gen", sed=TRUE)$sed^2 # obtain squared s.e.d. matrix
      	vdBLUP.avg <- mean(vdBLUP.mat[upper.tri(vdBLUP.mat, diag=FALSE)]) # take mean of upper triangle
      	heritabOutput$H2_Cullis[i] <- 1 - (vdBLUP.avg/(vc.g*2)) # H2 Cullis 
}

# rank the correlation and H2_cullis in descending order and sum the ranks
heritabOutput$rank_r <- rank(-heritabOutput$Corr)
heritabOutput$rank_H2 <- rank(-heritabOutput$H2_Cullis)
heritabOutput$sum_rank <- heritabOutput$rank_r + heritabOutput$rank_H2
heritabOutput <- heritabOutput[order(heritabOutput$sum_rank),]

cat("\n correlation and Cullis heritability estimate by  environments","\n")
print(heritabOutput)
write.csv(heritabOutput, file="output/fyld_H2_and_corr_per_env.csv", row.names=FALSE)

#cat("\n Blups estimate from individual environment","\n")
#print(BlupOutput)
  
# write the environment specific blups to csv file format
#write.csv(BlupOutput, file="output/fyld_env_specific_blup.csv")


#Plotting the number of representative environments
allEnvs <- unique(pheno$env)
nEnvs <- length(allEnvs)

## Cross validation
iters=50
ddList <- list()
for(u in 1:(nEnvs-1)){
   print(paste("nEnvs=",u))
  cors0 <- numeric()
  set.seed(1245)
   for(k in 1:iters){
     provENVS <- sample(allEnvs,u)
     provdat <- pheno[which(pheno$env %in% provENVS),]

     if(u==1){
	modfit  <- try(asreml(fixed=fyld~ prop_hav,
        random= ~ gen + at(env):rep,
        residual=~dsum(~units|env),
        na.action=na.method(y='include',x='include'),
        data=provdat,maxit=30,workspace = 250e6, pworkspace=250e6,trace=FALSE))


	if (!inherits(modfit, what = "try-error")){
		# extract out the predicted value from each trial and correlate with all env blups
        	PredVal <- predict(modfit, classify = "gen")$pvals
        	envPredVal <- PredVal %>% select(gen,predicted.value) %>%
        	rename(envBlup=predicted.value) %>% data.frame
        	mergePredVal  <- merge(BLUPlist_All, envPredVal, by="gen",all=TRUE)
		cors0[k] <- with(mergePredVal, cor(AllEnvBlup,envBlup, use="complete.obs"))
       	} else{
		cors0[k] <- NA
		}

} else{

	modfit  <-  try(asreml(fixed=fyld~prop_hav + env,
        	random= ~rr(env):gen +  diag(env):gen + at(env):rep,
        	residual=~dsum(~units|env),
        	na.action=na.method(y='include',x='include'),
       	 	data=provdat,maxit=30,workspace = 250e6, pworkspace=250e6,trace=FALSE))

	if (!inherits(modfit, what = "try-error")) {
	
		# extract out the predicted value from each trial and correlate with all env blups
                PredVal <- predict(modfit, classify = "gen")$pvals
                envPredVal <- PredVal %>% select(gen,predicted.value) %>%
                rename(envBlup=predicted.value) %>% data.frame
		mergePredVal  <- merge(BLUPlist_All, envPredVal, by="gen",all=TRUE)
                cors0[k] <- with(mergePredVal, cor(AllEnvBlup,envBlup, use="complete.obs"))
       }else{
		cors0[k] <- NA}
     }
   }
ddList[[u]] <- data.frame(r=cors0, iter=1:length(cors0), nENV=u)
}

cat("\n print ddlist","\n")
print(ddList)
envSamplingOutput <- do.call(rbind, ddList)

  gainOutput <- envSamplingOutput %>%
    group_by(nENV) %>%
    summarise(
      MEAN = mean(r, na.rm=TRUE),
      SE = sd(r)/sqrt(iters)#quantile(r,.95)
    )

# write out the gain from sampling environment representative
write.csv(gainOutput, file="output/fyld_env_sample_gainList.csv",row.names=FALSE)
```


# Representative of TPE for 3 traits
This is a line plot showing the trend of a number of environments likely to present target population of environment (TPE) for the traits: fyld, dm, and tyld.
Note: the output for individual traits is merged together i.e, row binds to have this plot.

```{r, TPE}

gainList <- read.csv(file="output/traitTPEGainList.csv", header=T)

TPEplot <- ggplot(gainList,aes(x=nENV,y=MEAN, group=trait,colour=trait))+
  geom_ribbon(aes(x=nENV,ymin=MEAN-SE,ymax=MEAN+SE),alpha=0.2,linetype=0)+
  geom_point()+
  geom_line(size=1)+
  guides(alpha=FALSE)+
  scale_x_continuous("nEnvironments", labels = seq(1,max(gainList$nENV),2), breaks = seq(1,max(gainList$nENV),2))+
  scale_y_continuous("Breeding value accuracy compared to all_Envs", limits=c(0,1), labels = seq(0,1,.1), breaks = seq(0,1,.1)) +
  geom_vline(xintercept=c(2,3,5,8), linetype="dashed", color = "red")


ggsave("output/TPEplot.jpeg",height=4.2, width=6.5, units="in", dpi=300)
print(TPEplot)
```



# Clustering testing environments based on genotypic BLUPs

```{r, eval=FALSE}

# fit joint analysis based of fresh root yield
modfit  <-  asreml(fixed=fyld ~env + prop_hav,
        random= ~ at(env):rep + gen + env:gen,
        residual=~dsum(~units|env),
        na.action=na.method(y='include',x='include'),
        data=pheno,maxit=150,workspace = 250e6, trace = FALSE)


cat("\n BLUPs by  environments","\n")
blupByEnv <- predict(modfit, classify = "env:gen",sed=T)$pvals

blupByEnv <- blupByEnv %>% dplyr::select(env,gen, predicted.value)%>%
                rename(predVal=predicted.value) %>% spread(env,predVal)

write.csv(blupByEnv, file="output/envSpecificBLUP.csv",row.names=FALSE) # write the BLUPs to a file
```


# Correlation among the environments based on the genotypic blups

```{r,clusterEnvs}
setwd("/Users/mab658/Documents/research_project/classical_analysis_GxE")
blupByEnv <- read.csv(file="../output/tyldenvSpecificBLUP.csv", header=T) 

# compute Pearson correlation among the environment-specific genotypic blups
corcoef <- cor(blupByEnv[,2:length(colnames(blupByEnv))], method = "pearson", use="pairwise.complete.obs")


# Coerce to matrix object
corcoef <- as.matrix(corcoef) # coerce to matrix object

# correlation distance matrix from correlation matrix
dist.mat <- as.dist(1-abs(corcoef))

# carry out hierarchical clustering based on the ward linkage method
res.hc <- hclust(d=dist.mat, method="ward.D2")

# plot(as.dendrogram(res.hc),ylab="Height",edgePar=list(col=3, lwd=3), cex=0.2, horiz=F) 
# Cut the tree

envplot <- fviz_dend(res.hc, cex = 0.6, k = 5, 
 color_labels_by_k = FALSE,  main="")

print(envplot)
ggsave("../output/envClusterBLUP_tyld_dendogram.jpeg",height=5, width=10, units="in", dpi=300)
```

